Data Specialization for a Staged Lambda Calculus

### 2015-02-26 paper/intro.tex

Such computations, where we apply a multivariate function to arguments
where one of the argument remains the same are common.  In fact, many
research problems aim at improving efficiency of such functions.  For
example, algorithm design techniques such as dynamic programming
identify the computation in $f$ that depend on $x$ try to ensure that
such computations are performed no more than once.  Similarly, crucial
compiler optimization techniques such as loop hoisting and common
subexpression elimination take identify common computations across all
invocations of $f$ and perform them once and for all at the beginning
of the computation. The whole area of incremental computation aims at
adapting a computation to small changes, which may be applicable to
this problem depending on the closeness of $y_i$'s.

The common theme in all such design and optimization techniques is to
achieve what Jorring and Scherlis back in 1980's called: {\em
  frequency reduction} or {\em precomputation}~\cite{JS86-staging}.
The idea behind frequency reduction is to identify computations that
are performed multiple times and pull them ahead so that they can be
performed once and used later as needed.  The idea behind
precomputation is to identify computations that can be performed
earlier, for example at compile time if their inputs are available
statically and perform them at that earlier time.

There has been much work on language-based techniques for performing
frequency reduction and precomputation.  Broadly speaking, these
techniques can be categorized into
\begin{itemize}
\item 
partial evaluation,

\item
meta programming, and

\item 
pass separation or data specialization.

\end{itemize}

Partial evaluation performs precomputation by distinguishing between a
{\em static} and {\em dynamic} stage and performs all static
computations first, followed be the dynamic stage.  For example,
suppose that we wish to compute $F(x, y_0 \ldots y_{m-1})$ for a
statically known $x = [~]$.  We can derive a specialization of $f$ at
$[~]$, written $f_{[~]}$, such that $f(x,y) = f_{[~]}(y)$ and specialize $F$
for $x=[~]$ as
\begin{lstlisting}
fun @$F_{[~]}(y_0, y_1 \ldots y_{m-1})$@ = 
  @$(f_{[~]}(y_0);  f_{0}(y_1)  \ldots f_{0}(y_{m-1}))$@.
\end{lstlisting}
%
Such a partial evaluation typically proceed by performing a
binding-time analysis to determine the static and dynamic parts of $f$
and $F$ and specializes them based on the results of the binding-time
analysis.

A closely related technique, {\em meta programming} enables the
programmer to write a program that takes the static values and
generates a specialized program for those values.  Meta programming
gives the programmer relatively fine-grained control over partial
evaluation by requiring staging annotations in the program.  Such
staging annotations make explicit the stage of each computation and
can overcome some of the intricacies of binding-time analysis but this
comes at the cost of more complex reasoning and some loss of
automation (partial evaluation can in principle be applied fully
automatically without relying on any annotations).

@@

While highly desirable, manual stratification is difficult and highly
nonmodular: program invariants are scattered over the code written in
multiple strata and because modification to any one stratum requires
modifying all others to match.  Ideally, we wish to write just one
program and stratify it into multiple strata automatically.  But such
automatic stratification turns out to be challenging: the first
successful approach that we know of is by Knoblock and
Ruf\,\cite{knoblock96}, which like our work was inspired by computer
graphics applications.  While practically effective, Knoblock and Ruf
consider only a very simple language and are not able to stratify
recursive programs.  Recent
efforts\,\cite{Proudfoot:2001,Foley:2011,He:2014} ....  but still
cannot stratify recursive programs.
%
In fact back in 80,s Jorring and Sherlis was able to stratify
reasonably complex programs but only manually and note that
``... partial evaluation technique is readily mechanized, while
staging transformation approach will elude full automation for some
time.''

In this paper we make significant progress on the problem of automatic
stratification of higher-order functional programs.  As our starting
point, we take the modal language \langname with explicit staging in
the style of Davies~\cite{Davies}, where the ``circle'' modality
denotes computation in a later stage, and present a staged operational
semantics similar for \langname (\secref{language}).  We then present
a splitting algorithm that stratifies a \langname program into a two
mono-staged programs, one for the first stage and for the second stage
(\secref{splitting}).  Since each program is at a mono-stage, they are
expressed in conventional function languages.

@@

Since it deals with unrestricted recursion and a rich language
consisting of first-class functions, recursion, and sum and recursive
types, the splitting algorithm is naturally complex. With some care,
the algorithm nevertheless can be specified reasonably succinctly.  In
fact, we have implemented the algorithm (\secref{implementation}) and
applied to several examples (\secref{examples}).  As an interesting
example, we consider a function
\begin{lstlisting}
fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
  @$f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@,
\end{lstlisting}
where $f(x,y)$ selects the element of the list $x$ with rank $y$ by
using the ``quickselect'' algorithm.  When applied to this function,
our split algorithm yields an asymptotically more efficient function
that runs in expected $O(n\log{n} + m\log{n})$ time instead of
expected $O(n \cdot m)$ time---a near linear time improvement.
Interestingly the code output by our splitting algorithm
auto-generates the ``quicksort'' algorithm, which it uses to build a
binary search tree as a boundary value in the first stage.  For the
second stage, the splitting algorithm generates a ``binary search''
that takes the tree and performs a binary search on the tree as guided
by the given rank.  To the best of our knowledge, no prior approaches
can perform such complex transformations on higher-order code, nor can
they yield such asymptotic improvements in run time.

>>>>

Such computations, where we apply a multivariate function to arguments
where one of the argument remains the same are common. In fact, many
research problems aim to improve the efficiency of such functions.  For
example, algorithm design techniques such as dynamic programming
identify the computation in $f$ that depend on $x$, and try to ensure that
such computations are performed no more than once.  Similarly, crucial
compiler optimization techniques such as loop hoisting and common
subexpression elimination identify common computations across all
invocations of $f$ and perform them once and for all at the beginning
of the computation. The whole area of incremental computation aims at
adapting a computation to small changes, which may be applicable to
this problem if the $y_i$'s are sufficiently similar.

The common theme in all such design and optimization techniques is to
achieve what in the 1980s J{\o}rring and Scherlis called {\em
frequency reduction} or {\em precomputation}~\cite{JS86-staging}.
The idea behind frequency reduction is to identify computations that
are performed multiple times and hoist them so that they can be
performed once and used later as needed.  The idea behind
precomputation is to identify computations that can be performed
earlier---for example, at compile time if their inputs are known
statically---and perform them at that earlier time.

There has been much work on language-based techniques for performing
frequency reduction and precomputation.  Broadly speaking, these
techniques can be categorized into
\begin{itemize}
\item partial evaluation,
\item metaprogramming, and
\item pass separation or data specialization.
\end{itemize}

Partial evaluation achieves precomputation by distinguishing between 
{\em static} and {\em dynamic} stages, and performing all static
computations first, followed by dynamic ones.  For example,
suppose that we wish to compute $F(x, y_0 \ldots y_{m-1})$ for a
statically known $x = [~]$.  We can derive a specialization of $f$ at
$[~]$, written $f_{[~]}$, such that $f(x,y) = f_{[~]}(y)$ and specialize $F$
for $x=[~]$ as
\begin{lstlisting}
fun @$F_{[~]}(y_0, y_1 \ldots y_{m-1})$@ = 
  @$(f_{[~]}(y_0);  f_{0}(y_1)  \ldots f_{0}(y_{m-1}))$@.
\end{lstlisting}
%
Such a partial evaluation typically proceed by performing a
binding-time analysis to determine the static and dynamic parts of $f$
and $F$ and specializes them based on the results of the binding-time
analysis.

A closely related technique, {\em metaprogramming}, enables the
programmer to write a program that takes the static values and
generates a specialized program for those values.  Metaprogramming
gives the programmer relatively fine-grained control over partial
evaluation by requiring staging annotations in the program.  Such
staging annotations make explicit the stage of each computation and
can overcome some of the intricacies of binding-time analysis but this
comes at the cost of more complex reasoning and some loss of
automation (partial evaluation can in principle be applied fully
automatically without relying on any annotations).

@@

While highly desirable, manual stratification is difficult and highly
nonmodular: program invariants are scattered over the code written in
multiple strata and because modification to any one stratum requires
modifying all others to match.  Ideally, we wish to write just one
program and stratify it into multiple strata automatically.  But such
automatic stratification turns out to be challenging: the first

Such automatic stratification turns out to be challenging: the first
successful approach that we know of is by Knoblock and
Ruf\,\cite{knoblock96}, which, like our work, was inspired by computer
graphics applications.  While practically effective, Knoblock and Ruf
consider only a very simple language and are not able to stratify
recursive programs.  Recent
efforts\,\cite{Proudfoot:2001,Foley:2011,He:2014} ....  but still
cannot stratify recursive programs.
%
In fact back in 80,s Jorring and Sherlis was able to stratify
reasonably complex programs but only manually and note that
``... partial evaluation technique is readily mechanized, while
staging transformation approach will elude full automation for some
time.''

In this paper we make significant progress on the problem of automatic
stratification of higher-order functional programs.  As our starting
point, we take the modal language \langname with explicit staging in
the style of Davies~\cite{davies96}, where the ``circle'' modality
denotes computation in a later stage, and present a staged operational
semantics similar for \lang\ (\ref{sec:stagedsemantics}).  We then
present a splitting algorithm that stratifies a \langname program into
a two mono-staged programs, one for the first stage and for the second
stage (\ref{sec:splitting}).  Since each program is monostaged, they
are expressed in conventional functional languages.

@@

Since it deals with unrestricted recursion and a rich language
consisting of first-class functions, recursion, and sum and recursive
types, the splitting algorithm is naturally complex. With some care,
the algorithm nevertheless can be specified reasonably succinctly.  In
fact, we have implemented the algorithm (\ref{sec:implementation}) and
applied to several examples (\ref{sec:examples}).  As an interesting
example, we show that a function

\begin{lstlisting}
fun @$F(x, y_0, y_1 \ldots y_{m-1}$@ = 
  @$f(x, y_0);  f(x, y_1)  \ldots f(x, y_{m-1})$@,
\end{lstlisting}
where $f(x,y)$ selects the element of the list $x$ with rank $y$ by
using the ``quickselect'' algorithm.  When applied to this function,
our split algorithm yields an asymptotically more efficient function
that runs in expected $O(n\log{n} + m\log{n})$ time instead of
expected $O(n \cdot m)$ time---a near linear time improvement.
Interestingly the code output by our splitting algorithm
auto-generates the ``quicksort'' algorithm, which it uses to build a
binary search tree as a boundary value in the first stage.  For the
second stage, the splitting algorithm generates a ``binary search''
that takes the tree and performs a binary search on the tree as guided
by the given rank.  To the best of our knowledge, no prior approaches
can perform such complex transformations on higher-order code, nor can
they yield such asymptotic improvements in run time.
